When does vapor pressure deficit drive or reduce evapotranspiration?

### 2017-12-13 doc/agu_2017/vpd.tex
refactoring slides

\section{Introduction}
\begin{frame}{Does VPD drive or reduce ET? - atmospheric demand perspective}
  \begin{Huge}
  \[VPD = (1-RH)\cdot e_s (T)\]
\end{Huge}
  \begin{itemize}
  \item Increase in VPD (\textbf{increase in atmospheric demand}) drives  an \textbf{increase in ET}.
  \end{itemize}
\end{frame}

\begin{frame}{Does VPD drive or reduce ET? - plant response perspective}
  \begin{columns}
    \begin{column}{0.5\textwidth}
      \begin{itemize}
      \item However, plants evolved to use stomata to conserve and regulate water use. So \textbf{stomata closure} in response to increases in VPD may \textbf{decrease ET}.
      \end{itemize}
    \end{column}
    \begin{column}{0.5\textwidth}
      \begin{figure}
        \includegraphics[width=2.25in]{stomata.png}% was O(1.5in)
        \caption{from evolution.berkeley.edu}
      \end{figure}
    \end{column}
  \end{columns}
\end{frame}

@@

\section{Method}
\begin{frame}{Develop a theory to quantify VPD effects}
  \begin{itemize}
  \item The goal is to simply the problem (increase transparency) while still capturing the leading order behavior the system. (intrinsic knowledge)
    \begin{itemize}
    \item While simplifying the system aids intrinsic understanding, simplifying assumptions sacrifice physical realism.
    \end{itemize}
  \item Because we run the risk of over-simplification, we will use FLUXNET2015 data to test how well our theory matches the data. 
  \end{itemize}
\end{frame}

\begin{frame}{Simple theory - start with Penman-Monteith}
  We can use Penman-Monteith (PM) to estimate ET:
  \[ET = \frac{\Delta R + g_a \rho_a c_p \, VPD}{\Delta + \gamma(1 + \frac{g_a}{\textcolor{magenta}{\boldsymbol{g_s}}})},\]
\textbf{Problem}: \Large $\textcolor{magenta}{\boldsymbol{g_s}}$ \normalsize (stomatal conductance) is a function of photosynthesis, which is a function of ET itself.  So ET in Penman-Monteith is really an implicit function of itself and we cannot take derivatives!
\end{frame}

@@

@@

@@

\begin{frame}{Summary}
  \small
  \begin{itemize}
  \item Derived a new form of Penman-Monteith that explicitly calculates ET given published constants.
  \item Developed a theory for when atmospheric dryness will increase ET (atmospheric demand dominates) vs. decrease ET (plant response dominates). Theory tests well using real data from FLUXNET2015 on plant types CSH, ENF, and DBF. Applicability to CRO and GRA is more uncertain.
  \item For plants evolved to prioritize water conservation over production, atmospheric demand is less likely to dominate.
  \item All plant types except for crops had common occurrence of decreases in ET with increases in VPD (plant response dominates). Therefore, the hydrometeorologists perspective is often wrong, and drought indices built upon assumptions about atmospheric demand are unrealistic for vegetated surfaces.
  \item Our form of Penman-Monteith is capable of estimating ET realistically and could be used as a replacement for PET in drought indices (e.g. Palmer Drought Severity Index - PDSI).
  \end{itemize}
\end{frame}

>>>>

\section{Introduction}

\begin{frame}{Does VPD drive or reduce ET? - atmospheric demand perspective}
  \begin{columns}
    \begin{column}{0.7\textwidth}
      Increase in VPD (\textbf{increase in atmospheric demand}) drives  an \textbf{increase in ET}.
      \[VPD = (1-RH)\cdot e_s (T)\]
    \end{column}
    \begin{column}{0.3\textwidth}<2>      
      \includegraphics[width=\textwidth]{towel.jpg}
    \end{column}
  \end{columns}
\end{frame}

\begin{frame}{Does VPD drive or reduce ET? - plant response perspective}
  \begin{columns}
    \begin{column}{0.5\textwidth}
      However, plants evolved to use stomata to conserve and regulate water use. So \textbf{stomata closure} in response to increases in VPD may \textbf{decrease ET}.
    \end{column}
    \begin{column}{0.5\textwidth}
      \begin{figure}
        \includegraphics[width=2.25in]{stomata.png}% was O(1.5in)
        \caption{from evolution.berkeley.edu}
      \end{figure}
    \end{column}
  \end{columns}
\end{frame}

@@

\section{Method}
\begin{frame}{Develop a theory to quantify VPD effects}
  \begin{itemize}
  \item The goal is to simply the problem (increase transparency) while still capturing the leading order behavior the system.
    \begin{itemize}
    \item While simplifying the system aids intrinsic understanding, simplifying assumptions sacrifice physical realism.
    \end{itemize}
  \item Because we run the risk of over-simplification, we will use FLUXNET2015 data to test how well our theory matches the data. 
  \end{itemize}
\end{frame}

\begin{frame}{Simple theory - start with Penman-Monteith}
  We can use Penman-Monteith (PM) to estimate ET:
    \[ET = \frac{\Delta R + g_a \rho_a c_p \, VPD}{\Delta + \gamma(1 + \frac{g_a}{\textcolor{magenta}{\boldsymbol{g_s}}})},\]
  \begin{overprint}
    \onslide<2>
    \textbf{Problem}: \Large $\textcolor{magenta}{\boldsymbol{g_s}}$ \normalsize (stomatal conductance) is a function of photosynthesis, which is a function of ET itself.  So ET in Penman-Monteith is really an implicit function of itself and we cannot take derivatives!<2>
  \end{overprint}
\end{frame}

@@

\begin{frame}{Consequences of theory - ``sign'' term}
  \[\frac{\partial \, ET}{\partial \, VPD} = \text{scaling} \cdot \left(\frac{ c_p}{R_{air}} - \frac{\gamma c_s }{1.6 \; R*\; \text{ uWUE }} \left( \frac{2 g1 + \sqrt{VPD}}{2 (g1 + \sqrt{VPD})^2}\right) \right)\]
  \begin{itemize}
  \item $c_p$ and $R*$ are constants
  \item $R_{air}$, $\gamma$, and $c_s$ are approximately constant (relative to $\sqrt{VPD}$)
  \item uWUE and $g1$ are constants within plant type (e.g. grass, crops, deciduous broadleaf forest, evergreen needleleaf forest, shrub)
  \end{itemize}
  So \textbf{within each plant type}, whether the atmospheric demand (ET increasing with VPD) or plant response (ET decreasing with VPD) dominates is approximately \textbf{just a function of VPD}!
\end{frame}

@@

\begin{frame}{Test theory with FLUXNET data - the good}
  \begin{columns}
    \begin{column}{0.33\textwidth}
      \includegraphics[width=1.2\textwidth]{DBF_box.pdf}
    \end{column}
    \begin{column}{0.33\textwidth}
    \end{column}
    \begin{column}{0.33\textwidth}
    \end{column}
  \end{columns}
\end{frame}

@@

\begin{frame}{Summary - When does VPD drive or reduce ET?}
  \begin{itemize}
  \item Theory predicts that each plant type has a \textbf{critical VPD} \textbf{below which ET will decrease} (plant response dominates), and \textbf{above which ET will increase} (atmosheric demand dominsates).
  \item For forest sites, environmental VPD approximately straddles the critical VPD.
  \item For shrubs environmental VPD never exceeds the critical VPD.
  \item Theory tested poorly with FLUXNET data for for crops and grass. 
  \item All plant types exhibited a response far below that of PET.
  \item The new uWUE-version of Penman-Monteith we derived could be used as a replacement for PET in drought indices over vegetated surfaces.
  \end{itemize}
\end{frame}
